Efficient Parallelization of $\ell_1$-Minimization for Face Recognition \thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}

### 2011-02-02 paper/main.tex
Reflowed some sections, Re-organized a bit.

\subsection{Computational Complexity}
Any discussion of the speed of an algorithm should discuss the comutational complexity of the algorithm.  For the ALM
algorithms considered in this paper, this is rather simple.
The Matrix-vector computions are $O(mn)$, and the vector-vector operations are $O(m)$, 
in terms of both computation cost and memory-bandwidth.  Since A is very tall, 
the computation of $A^\dagger$, the pseudoinverse of $A$, is best computed via $A^\dagger = (A^TA)^-1 A^T$.
The complexity of computing $G = A^T A$ is $O(n*n*m)$.  $G$ is sufficiently small that it can be inverted via
Gauss-Jordan elimination, which has a complexity of $O(n^3)$.  The computation of $A$ finishes with a matrix
multiplication, with cost $O(n^2 m)$.
Fortunately, $A^\dagger$ only needs to be computed once per instance of the $\ell_1$ minimization problem.

@@

\section{CPU Implementation}
\subsection{CPU Hardware Parallelism}
The baseline architecture for our experiments is 
a typical high end workstation with two quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.
Each processor has its own memory interface, and is directly connected to half of the
RAM installed in the machine.  The amount of RAM installed exceeds the amount used
by the algorithms, and is not an important performance consideration.

Each core has a 32 KB L1 data cache and a 256 KB L2 cache. With Hyperthreading enabled, each core can
support two threads simultaneously.  For floating point math, each core also has a vector unit capable of
performing the same arithmetic operation on 4 single precision floating point values simultaneously.
\footnote{The core also has scalar floating point units, but some compilers, notably gcc, generate
code that uses the vector units exclusively for floating point math.}

@@

Each processor has 8MB of L3 cache that is shared by all 4 cores.
Overall, this means the algorithm has approximately 16MB of on-chip cache available.  
For the low resolution images used in sparse representation based face recognition,
this is enough cache to hold over 800 image-sized variables, meaning that for small data sets
it is possible to run completely using on-chip cache.

%\subsection{Tuning the GEMV operation}
%When matrix-vector operations are called sequentially on the same data,
%the operation can be blocked so that each thread operates on the same block of data for each operation.
%For Single threaded MKL sgemv, Code took          0.0335823 msec per instance to run!
%For Automatic multi-threaded MKL sgemv, Code took 0.008179 msec per instance to run!
%For Manually multi-threaded MKL sgemv, Code took  0.0061276 msec per instance to run!

%\subsection{Tuning the Soft-Thresholding Operator}
%...Make sure there are no branches
%...Make sure there are no function calls
%...Make sure the loop gets automatically vectorized
%...Block the computation into work for different cpu cores

\section{GPU Implementation}
\subsection{GPU Hardware Parallelism}
High-level description of GPU parallelism, i.e.

@@

%You should give a brief (1-2 paragraphs) overview of the GPU architecture -- most people have heard of GPUs and what they can do by now, so it isn't worth spending a whole lot of space in the paper describing it.

%You should define the levels of the Cuda thread and memory hierarchies: i.e. warps, thread blocks, grids and register file, shared memory, global memory. Also mention that the GPU memories are all distinct from the CPU memories, and data transfers go over PCI-Express -- but the cost of those transfers are amortized by doing lots of work (i.e. solving an $\ell_1$ minimization after transferring the matrix once). It is worth spending a little space describing how the GPU is a multi-core processor (just like the CPUs), but with much more sophisticated vector processing hardware. That is: on a Fermi there are 15 cores, and each one has two 16-wide vector floating-point units.

%The GPU-algorithm is the same as the CPU algorithm: primary- and dual- ALM methods. The implementations differ in that you have to parallelize the GPU code at an additional level, and you can't use the GPU's caches to communicate between cores (only within a core). On the CPU, you only need to parallelize for the 4 or 8 cores -- and perhaps add SSE instructions when possible (usually this is inside of the BLAS library) -- any communication between cores happens implicitly via cache coherence. For the GPU, you need to parallelize for the 15 cores, but also for the 32-way SIMD parallelism of Warps. Any intra-core communication happens via the cache (aka scratchpad, or __shared__ memory) and within-block barriers (i.e. __syncthreads()), and any inter-core communication happens via the GPU's DRAM and the implicit barriers between grid launches.

cuda cores, SIMD width, scheduling restrictions, bandwidth
\subsection{Solving a Single $\ell_1$ minimization}

@@

\section{Benchmarks on a Single Large $\ell_1$ Problem}

In this section, we present $2$ sets of experiment to benchmark the performance
of the PALM algorithm on CPU and GPU platforms. 
The first set of experiments will compare implementations that solve a single
large instance of an $ell_1$ minimization problem.  
The second set of experiments will compare implementations that must solve
many instances of $\ell_1$ minimization.

To generate a meaningful comparison between different hardware architectures,
in all cases we compare the performance of hardware architectures on a
per-board basis:  For CPU implementations, the benchmark makes use of all of
the cores in as many cpu's are present.  For GPU implementations, the benchmark
makes the best use of the entire GPU chip (most GPU boards have a single GPU
chip).

Unless otherwise specified, all implementations utilize single precision
floating point datatypes.
% and were verified to output results differing by no
%more than $10^{-4}$.

@@

\subsection{PALM $\ell_1$ minimization benchmark on Synthetic Data}

In this section, we benchmark the performance of MATLAB, BLAS MKL, and GPU CUDA
implementations of the Primal ALM algorithm on synthetic data to find at which
point the GPU implementation runs faster than CPU implementation.  To simply
the benchmark, we cap the number of inner iterations at 50 and outer iterations
at 5000 and run the algorithm to convergence.  CPU code was run on an Intel
Dual Socket Quad Core with 8 gigs of DDR3 RAM.  GPU code was run on a GTX480 We
ran the A matrix with different ratios between m:n from 10:1:5 for
m:2000:25:8000 with an x vector of 10 percent sparsity.  Tolerance levels were
empirically seen to allow best performance with the inner loop tolerance set at
one order of magnitude lower than the outer loop tolerance.

@@

%An example of image inclusion for Victor:
\begin{figure*}
\centering
\includegraphics[width=3.5in]{figures/PALM_benchmark_ratio_1.png}
\caption{This is where the caption goes.}
\label{fig:uniqueidentifierforthisimage}
\end{figure*}

% change picture to only include CPU/GPU not MATLAB

As the dimensions of A increase, so does the significance of the size of AtA.  Until the A and AtA matrix reach x Mb, the CPU version runs faster than the GPU.

\section{Benchmarks on Solving Many $\ell_1$ Problems Concurrently}

%This section is about Drew's kernels for solving multiple $\ell_1$ minimizations in parllel

%How to efficiently perform an $\ell_1$ minimization on a single core, quote performance on a large batch.

%How to solve multiple instances in smaller batches without losing concurrency, using CUDA streams.

%...Mention effect of compiler verion on streams implementation

\section{Conclusion}
...How lessons from this paper extend (or don't extend) to other similar problems in computer vision

...Reiterate how the appropriate hardware choice depends on the problem size.

>>>>

\subsection{Computational Complexity} Any discussion of the speed of an
algorithm should discuss the comutational complexity of the algorithm.  For the
ALM algorithms considered in this paper, this is rather simple.  The
Matrix-vector computions are $O(mn)$, and the vector-vector operations are
$O(m)$, in terms of both computation cost and memory-bandwidth.  Since A is
very tall, the computation of $A^\dagger$, the pseudoinverse of $A$, is best
computed via $A^\dagger = (A^TA)^{-1} A^T$.  The complexity of computing $G =
A^T A$ is $O(nnm)$.  $G$ is sufficiently small that it can be inverted via
Gauss-Jordan elimination, which has a complexity of $O(n^3)$.  The computation
of $A$ finishes with a matrix multiplication, with cost $O(n^2 m)$.
Fortunately, $A^\dagger$ only needs to be computed once per instance of the
$\ell_1$ minimization problem.

@@

\section{Hardware Parallelism}

\subsection{CPU Hardware Parallelism} 
The baseline architecture for our experiments is a typical workstation,
with two quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each
processor has its own memory interface, and is directly connected to half of
the RAM installed in the machine.  The amount of RAM installed exceeds the
amount used by the algorithms, and is not an important performance
consideration.

Each core has a 32 KB L1 data cache and a 256 KB L2 cache. With Hyperthreading
enabled, each core can support two threads simultaneously.  For floating point
math, each core also has a vector unit capable of performing the same
arithmetic operation on four single precision floating point values
simultaneously.  \footnote{The core also has scalar floating point units, but
some compilers, notably gcc, generate code that uses the vector units
exclusively for floating point math.}

@@

Each processor has 8MB of L3 cache that is shared by all four cores.  Overall,
this means the algorithm has approximately 16MB of on-chip cache available.
For the low resolution images used in sparse representation based face
recognition, this is enough cache to hold over 800 image-sized variables,
meaning that for small data sets it is possible to run completely using on-chip
cache.

\subsection{GPU Hardware Parallelism}

@@

%You should define the levels of the Cuda thread and memory hierarchies: i.e.
%warps, thread blocks, grids and register file, shared memory, global memory.
%Also mention that the GPU memories are all distinct from the CPU memories, and
%data transfers go over PCI-Express -- but the cost of those transfers are
%amortized by doing lots of work (i.e. solving an $\ell_1$ minimization after
%transferring the matrix once). It is worth spending a little space describing
%how the GPU is a multi-core processor (just like the CPUs), but with much more
%sophisticated vector processing hardware. That is: on a Fermi there are 15
%cores, and each one has two 16-wide vector floating-point units.

%The GPU-algorithm is the same as the CPU algorithm: primary- and dual- ALM
%methods. The implementations differ in that you have to parallelize the GPU
%code at an additional level, and you can't use the GPU's caches to communicate
%between cores (only within a core). On the CPU, you only need to parallelize
%for the 4 or 8 cores -- and perhaps add SSE instructions when possible
%(usually this is inside of the BLAS library) -- any communication between
%cores happens implicitly via cache coherence. For the GPU, you need to
%parallelize for the 15 cores, but also for the 32-way SIMD parallelism of
%Warps. Any intra-core communication happens via the cache (aka scratchpad, or
%__shared__ memory) and within-block barriers (i.e. __syncthreads()), and any
%inter-core communication happens via the GPU's DRAM and the implicit barriers
%between grid launches.

% In the followingn 2 sections, just describe implementations; single number
% benchmarks and microbenchmarks are ok, but save big graphs for experiments
% section.

% This section is Victor's 
\section{Single Instance Implementations}

\subsection{Single Instance CPU Implementations}

\subsection{Single Instance GPU Implementations}

@@

% This section is Drew's
\section{Multiple Instance Implementations}

\subsection{Multiple Instance CPU Implementations}
%\subsection{Tuning the GEMV operation}
%When matrix-vector operations are called sequentially on the same data,
%the operation can be blocked so that each thread operates on the same block of data for each operation.
%For Single threaded MKL sgemv, Code took          0.0335823 msec per instance to run!
%For Automatic multi-threaded MKL sgemv, Code took 0.008179 msec per instance to run!
%For Manually multi-threaded MKL sgemv, Code took  0.0061276 msec per instance to run!

%\subsection{Tuning the Soft-Thresholding Operator}
%...Make sure there are no branches
%...Make sure there are no function calls
%...Make sure the loop gets automatically vectorized
%...Block the computation into work for different cpu cores
\subsection{Multiple Instance GPU Implementations}

% Describingn the sequence of GPU implementations Drew has tried...

\section{Benchmarks}
In this section, we present two sets of experiment to benchmark the performance
of the PALM algorithm on CPU and GPU platforms. 
The first set of experiments will compare implementations that solve a single
large instance of an $ell_1$ minimization problem.  
The second set of experiments will compare implementations that must solve
many instances of $\ell_1$ minimization.

To generate a meaningful comparison between different hardware architectures,
in all cases we compare the performance of hardware architectures on a
per-board basis:  For CPU implementations, the benchmark makes use of all of
the cores in as many cpu's are present.  For GPU implementations, the benchmark
makes the best use of the entire GPU chip (most GPU boards have a single GPU
chip).  Unless otherwise specified, all implementations utilize single
precision floating point datatypes.

\subsection{Single Large $\ell_1$ Problem, Synthetic Data}
In this section, we benchmark the performance of MATLAB, BLAS MKL, and GPU CUDA
implementations of the Primal ALM algorithm on synthetic data to find at which
point the GPU implementation runs faster than CPU implementation.

@@

To simply the benchmark, we cap the number of inner iterations at 50 and outer
iterations at 5000 and run the algorithm to convergence.  CPU code was run on
an Intel Dual Socket Quad Core with 8 gigs of DDR3 RAM.  GPU code was run on a
GTX480 We ran the A matrix with different ratios between m:n from 10:1:5 for
m:2000:25:8000 with an x vector of 10 percent sparsity.  Tolerance levels were
empirically seen to allow best performance with the inner loop tolerance set at
one order of magnitude lower than the outer loop tolerance.

@@

\begin{figure*}
\centering
\includegraphics[width=3.5in]{figures/PALM_benchmark_ratio_1.png}
\caption{This is where the caption goes.}
\label{fig:uniqueidentifierforthisimage}
\end{figure*}

As the dimensions of $A$ increase, so does the significance of the size of
$A^TA$.  Until the $A$ and $A^TA$ reach ???? Mb (SHOULD BE MB?), the CPU version runs faster
than the GPU.

\subsection{Multiple Small $\ell_1$ Problems, Synthetic Data}

\subsection{Alignment Stage, Real Face Data}

\section{Conclusion}
